Efficiency

### 2018-01-29 DraftExec.tex

If it was known that the reactor couldn't be restarted after the poison-out period, the mission would need to be aborted in any case, and some kind of rescue sought. Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine was literally at the entrance of a home port.  It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a seach/rescue mission to rescure the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.

Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.  Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.  The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramtic reductions, will be dominant considerations in a nuclear submarine's power output.  To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumeably near-catastrophic) emergency.

@@

Most propellers will have between three and seven blades.  In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.  The blades tend to be twised so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.  The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement throug the water (rake). Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.

@@

\subsubsection{Pitch}
Perhaps one of the most important characteristics of a propeller is its pitch.  The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.  It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.  Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unitless equivalent, as is common in the description of many characteristics of propellers and propulsion systems.

@@

The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendancy for the blades to twist.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.  It is for this reason that a considerable number of vessels actually have variable pitch propellers.  Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.  Controllable pitch propellers also have advangates in terms of manouverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.  This has advantages for ferries which undertake frequent docking manouvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).

@@

Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.  Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propoulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.  Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio. In practice however, under self-propuslion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}

@@

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requieres the thust to be excerted over some distance travelled.

@@

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an acelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. \footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}

@@

A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved. In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.  In the decellerating case, the available area contracts after the water has passed the impeller. In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.

@@

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendancy for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particulary type of tubulence, and the onset of this type of cavitation, or the vortices that occur even when cavitation does not, can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accleratint duct, the propensity for cavitation other surface of the propeller is generally increased.  In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.

@@

A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}. A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.  A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low waterspeeds.  At higher waterspeeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.  When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.  Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).

@@

It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.  In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller". Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.  Other authors give separate treatement to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream sufrace that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaing a larger chord lenght right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which woud not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

@@

The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of diffeent pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

@@

The contrasting system is that of a radial or centrifugal pump.  In one of these systems, work is done on the water by spinning outwards, from close to the axis of rotation to the extremity. In such a case, the axis of the rotor is actually perpendicular to the direction of flow of the water exiting the pump.  The pumps are generally favoured for pump systems which need to add quite a large amount of head, mostly in the form of pressure, to a given amount of flow.  Consequently they are desirable for fire pumps, and or other systems which have to raise water to a high pressure, including lifting water up to great heights. However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in ouboard waterjets \parencite{yamaha2018}.

@@

The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accomodate this type of change.  However, the core limitations and physics surrounding duct design before and after the impeller remain in place for mixed-flow pumpjets, with the primary additional consideration being that the pump section of the duct will tend to need to be somewhat more bulbous in shape to accommodate the larger hub.  As such, differences in the pump efficiency (how well the impellers add energy to the flow) at any particular flow rate might not represent a large impact on the overall efficiency of the propulsor, as the losses due intake duct and nozzle will comprise separate and independent considerations, quite independent of the efficiency of the pump design.

@@

\subsection{Modern literature and results}
Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.  It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.  Fewer still tend to express their work in terms of ship-speed, instead adopting the convetion of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.  However, from a few notable studies exist which provide important reference points. (Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)

@@

Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.  They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.  However, care must be taken making precise quantitative comparisons, since such experiements involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar. Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.  Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.

At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particulalry for an anti-submarine warfare role. The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modelled and also tesed in demonstrator vessels.  These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.  Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focussing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal waterspeed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hullform in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impcat on the resistance offerd by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hullform needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.  As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.

@@

Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.  This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.  To more firmly establish the extent to which adaptation or plausible improvments in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.  This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.

@@

\begin{figure}[h]
\captionsetup{width=.8\textwidth}
\includegraphics[width=.8\textwidth]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}

>>>>

If it was known that the reactor couldn't be restarted after the poison-out period, the mission would need to be aborted in any case, and some kind of rescue sought. Even a hundred of tonnes of batteries embarked wouldn't change this reality, unless the submarine was literally at the entrance of a home port.  It seems much more likely that the safety requirement for batteries on a nuclear submarine is most likely to be framed around maintaining key life support systems (CO2 scrubbing for example) for sufficient time to enable some hope for a search/rescue mission to rescue the crew, in the likely case that the kind of emergency which shut down the reactor was the kind of emergency which meant the submarine might not resurface, such as the case of the Russian nuclear submarine, the Kursk.

Claims that the power management imperatives of nuclear submarines and diesel electric submarines are even remotely similar at low speeds are generally implausible.  Nuclear submarines will have unconstrained energy reserves, and a vast surplus of spare power capacity when ever operating below top speed.  The dynamics of reactor management dictate that generally avoiding large fluctuation in power, and strictly avoiding sharp and dramatic reductions, will be dominant considerations in a nuclear submarine's power output.  To put things simply, the characterisation that nuclear submarines have "power to burn" is generally sound, and would only ever be challenged when the submarine is travelling at flank speed, or the reactor is shut down due to a (presumably near-catastrophic) emergency.

@@

Most propellers will have between three and seven blades.  In general they are shaped as an aerofoil, with the convex side being upstream, just as the convex side of a plane wing is above, in order to generate lower pressure and lift as it moves through the air.  The blades tend to be twisted so that they have a higher angle of attack closer to the hub, and lesser closer to the extremities, so that those faster moving sections push their respective parts of the water column at an overall similar speed.  The are often also thinner towards the extremities, and will be swept backwards as if dragged by their rotation through the water (skew) and also dragged by the ship's movement through the water (rake). Whilst the degree of all these characteristics is highly variable for different applications, these are a few of the commonly referred to characteristics which can be varied in order to optimise performance for any given application.

@@

\subsubsection{Pitch}
Perhaps one of the most important characteristics of a propeller is its pitch.  The pitch represents the distance that a blade section would travel forward if it carved its way around a helix through one full rotation.  It is intuitive to think of as something akin to the angle-of-attack of the propeller blade to the water, however this is technically misleading as the movement of the water-column incoming to the blade, as well as the speed of rotation, also have a significant bearing on what the actual angle of attack of the blade ends up being.  Whilst technically pitch is a distance (measured in meters), it is often expressed and used in formula as a ratio of the diameter of the propeller, to give it a unit-less equivalent, as is common in the description of many characteristics of propellers and propulsion systems.

@@

The selection of the exact blade selection that is used to define pitch must be specified as being at some fraction of the radius from the centre, given the tendency for the blades to twist.  This distance is calculated to be the 'moment mean' or a technical derived effective average, which tends to lie between 0.6R and 0.7R.  A thorough discussion of pitch can be found in \cite[35-37]{carlton2007}.

Pitch is of particular importance to the discussion the efficiency of propeller and propulsor design because its optimal choice tends to vary considerably with the different loads which a vessel is intended to operate at, which can also relate to a vessel's design speed.  It is for this reason that a considerable number of vessels actually have variable pitch propellers.  Whilst considerably more complex, expensive, and heavier than a traditional fixed-pitch propeller, the ability to vary the pitch of a propeller assists considerably, particularly when the amount of load (resistance) a vessel is expected to face varies considerably.  Controllable pitch propellers also have advantages in terms of manoeuverability, since the pitch can be reversed and hence reverse thrust can be produced, without requiring the direction of the power coming from the engines to be reversed.  This has advantages for ferries which undertake frequent docking manoeuvres in confined spaces (\cite{MAN2017}, \cite{lewis1988}).

@@

Where $V_a$ is the speed of advance of the propeller, or the speed at which water arrives at the front-side of the propeller, and $n$ represents the rate of the propeller's rotation (revolutions per second) and $D$ is the propeller's diameter.  For fullness of understanding it is also worthwhile to note here that $V_a$ does not correspond directly to the vessel's speed of travel, though they are related. To some extent, a vessel moving through water will always draw water in its wake along with it, which means that in the wake (where the propeller tends to operate to some extent or other) the water incident on the propeller will arrive at a lower speed than that implied by the ship speed \parencite[15]{MAN2017}.  Another key distinction for the consideration of jets later on is that the speed of advance does not necessarily incorporate the movement of water due to the movement induced by the propeller (or duct) itself, which is a distinct effect to that of the wake field. In \cite[213]{lewis1988} $V_A$ is used specifically to denote the undisturbed stream velocity, since the role of certain ducts can be to accelerate or decelerate the stream prior to the water arriving at the blades of the propeller.

One may also note from the previous figure that for each line representing a given pitch, the efficiency falls precipitously towards zero around the point where $J$ reaches $P/D$. This is because this extreme of the curve represents a circumstance where the propeller is rotating, but not actually moving any water, since it is moving through the water at about that speed which corresponds to the exact helical path defined by the pitch being traced. This represents circumstances where a propeller encounters effectively zero load.  This might be the case when a eases it's power, and the momentum of the ship temporarily keeps the ship moving without the propeller exerting much or any net thrust, or where alternative propulsion sources (sails, or other propellers) are keeping a speed higher than what the propeller actually has push backwards on water to maintain.  Since drag due to friction scales with velocity squared, very low speed do correspond to increasing advance ratios for a given propeller size, as would increases in propeller size for a given speed and pitch ratio. In practice however, under self-propulsion conditions for open propellers, slowing down towards does not elevate the advance ratio far enough past beyond peak efficiencies for overall efficiency to fall very far, as can be seen in Figure \ref{fig:EfficiencyBulten.png}

@@

It can be seen from the diagram of the open-water efficiencies that there is some desirable non-zero slip ratio for a propeller.  The other end of the spectrum, where efficiencies also fall off toward zero, represent circumstances where the propeller is simply spinning in the same place in the water. In this case the slip ratio is one.  This represents a circumstance known as 'bollard pull' where a vessel is simply exerting force against an immovable object or infinite load.  Whilst the very low efficiencies here could represent a case of a propeller that is simply too small for the task it is trying to achieve, it is actually a significant and important circumstance to consider for propellers on working vessels such as tugs or trawlers, which exert large forces at very low water-speeds.  Measurement of the thrust generated in bollard pull, and the power required to achieve this thrust, is still an important metric for such vessels, but the definition of 'useful work' for normal propulsion requires the thrust to be exerted over some distance traveled.

@@

A detailed description of the essential differences between the two can be easily demonstrated by diagram, which show a simple comparison an accelerating and decelerating duct, which have the shape of the aerofoil section which comprises the duct inverted. \footnote{It should be noted that some designs attempt to strike a balance between efficiency and cavitation performance of accelerating and decelerating ducts for certain purposes, and in such cases the duct shape might not be so clearly contrasted \parencite{abdel2010}.  But for the purposes of explanation, as well as optimisation for military purposes, the distinction between the two will generally be quite clear, and in practice ducts will still tend to decelerate the flow on approach to the impeller unless specifically shaped to narrow in the opening.}

@@

A consideration of what happens to the flow as it moves through each type of duct helps to demonstrate how the acceleration and deceleration effects are achieved. In the case of an accelerating nozzle, the area through which the flow must pass contracts prior to the water arriving at the impeller.  In the decelerating case, the available area contracts after the water has passed the impeller. In some cases, a degree of diffusion is actually incorporated ahead of the impeller, whereby the area actually of the flow actually expands ahead of the impeller, which slows down the flow and increases its pressure according to Benoulli's principal \parencite{wislicenus1973}.

@@

A further means by which the efficiency of a propeller can be increased is through the reduction of the turbulence created by the blade-tips passing through the open (static) water at high speed.  Since the two sides of the blades have low and high pressure on them respectively, at the tips where the two faces meet, there there is a tendency for water to 'spill' from the high pressure side the the low pressure side, creating a vortex which leads to unwanted turbulence.  It is also around this vortex that blade tip cavitation tends to occur. By limiting the movement of water around the outside of the blade tip by having a tightly fitted shroud, losses due to this particular type of turbulence, and the onset of this type of cavitation, or the cortices that occur even when cavitation does not, can be diminished.

However, it is important to note that the existence of an accelerating duct by no means eliminates all cavitation.  In fact, given that the static pressure at the impeller is necessarily reduced by an accelerating duct, the propensity for cavitation other surface of the propeller is generally increased.  In many circumstance cavitation can and does still occur in ducted propellers near the blade tips at design loads \parencite{moulijn2015}.

@@

A detailed explanation of the means by which the ideal efficiency of propeller can be improved can be found in \cite[213-222]{lewis1988}, as well as other texts on marine propulsion, such as \cite{carlton2007}. A substantial work exploring the efficiencies of a range of accelerating and decelerating ducts can also be found in \cite{oosterveld1970}.  A common feature of these extensive studies also show that for ducted propellers, including accelerating ducted propellers, the advantage in terms of efficiency is overall restricted to low water-speeds.  At higher water-speeds, the drag induced by the water-movement over the exterior of the duct begins to increase substantially, and the net thrust on the duct becomes negative.  When the exterior of the duct experiences relatively little water velocity, but the interior experiences a great deal more, the greatest advantages of an accelerating duct are realised.  Furthermore, in general these models do not incorporate the effect of flow separation, which is acknowledged as a possibility on both the exterior and interior of ducts, which can dramatically reduce efficiency performance (\cite[20]{oosterveld1970}, \cite[214]{lewis1988}).

@@

It is perhaps most relevant at this point to further clarify the distinction between the different names for related or similar systems.  In \cite[17]{carlton2007} pump jet is described as an "interesting development of the classical ducted propeller". Oosterveld, in his thesis on ducted propellers also states with regard to the decelerating duct "This ducted propeller system is the so called pumpjet" \parencite[8]{oosterveld1970}.  Other authors give separate treatment to ducted propellers and pump jets, though the systems they describe are essentially similar in terms of composition, advantages, and physics \parencite[288]{lewis1988}.

@@

In other definitive literature on the design of pumpjets \parencite{henderson1964} the difference between a pumpjet and a propeller is a described as being that "the stream of flow through the pumpjet is made to depart from the "natural" or free-stream surface that bounds the flow though a standard propeller."  According to the diagrams and the subsequent description it is clear that the definition embraces quite exactly the distinction between an accelerating duct, and a decelerating duct, namely that the flow is slowed at the point it encounters the impeller, where as in a free situation the free stream would contract and accelerate at this point.

The addition of some new parts such as stator blades (to straighten out the flow) and other possible features which make the propeller system more like pump or turbomachinery components might be associated with systems more commonly referred to as pumpjets than ducted propellers.  Not all decelerating ducts have stators, but the impact of the duct on the stream will remain. The key concepts behind the operation of these systems, however, remains essentially similar in their nature and design intent.  In the presence of a duct propellers might be advantageously shaped differently, including maintaining a larger chord length right to the shroud, rather than tapering as open propellers tend to in order to lessen vortices and cavitation on the extremities.  However, there is not to my knowledge any single defining physical characteristic which the transition to the name 'impeller' implies, which would not be effectively equivalent to specialised propellers for decelerating ducts.  In systems tending to be described as pumpjets or waterjets hubs tend to be larger, blade-numbers higher, twist, skew and rake lower.  No definitive transition point seems to be accepted as having great significance.  Most commonly literature will refer to a 'rotor', a more general term which could encompass both propellers and impellers.

@@

The use of the word waterjet, as opposed to pumpjet, has a tendency signify that the jet expelled from the shroud or pump is situated above the water, which is desirable in very fast-moving surface vessels which attempt to minimise their contact with the waters surface, including high-speed planing hull vessels, including pleasure craft.  The physics of the two systems, however, are closely related, with the very noticeable exception of the absence of flow considerations for water on the exterior of the duct considerations for waterjets.  However, the physics relating to the impeller and stator blades or (turbomachinery) designed to add energy to the flow are generally extremely closely linked, though a larger variety of different pump types can be chosen for waterjets which might differ more significantly from the natural evolution of a propeller to an axial-flow pump. Many of the essential equations which govern the efficiencies and cavitation performance of pumpjets and waterjets are of the same or extremely similar form, as can be seen in detailed discussions given to both, by similar authors (\cite{wislicenus1973}, \cite{henderson1964}).  There are, however, plenty of instances in recent literature where this distinction regarding the ejection of the jet above and below the waterline is not maintained, and waterjets and pumpjets seem to be used interchangeably \parencite{abdel2010}, or systems are described as 'submerged waterjets' \parencite{buckingham2008submarine}.

@@

The contrasting system is that of a radial or centrifugal pump.  In one of these systems, work is done on the water by spinning outwards, from close to the axis of rotation to the extremity. In such a case, the axis of the rotor is actually perpendicular to the direction of flow of the water exiting the pump.  The pumps are generally favoured for pump systems which need to add quite a large amount of head, mostly in the form of pressure, to a given amount of flow.  Consequently they are desirable for fire pumps, and or other systems which have to raise water to a high pressure, including lifting water up to great heights. However, centrifugal pumps were originally the type of pump used in the first waterjets created by Hamilton \parencite{hamilton1997}, and are still used in some waterjet applications today, for example in outboard waterjets \parencite{yamaha2018}.

@@

The physics of of the fluid flows through the impeller of a mixed-flow pump are quite different to that of a pure-axial pump (the flows must be modeled in three rather than two dimensions), and blade design necessarily changes somewhat to accommodate this type of change.  However, the core limitations and physics surrounding duct design before and after the impeller remain in place for mixed-flow pumpjets, with the primary additional consideration being that the pump section of the duct will tend to need to be somewhat more bulbous in shape to accommodate the larger hub.  As such, differences in the pump efficiency (how well the impellers add energy to the flow) at any particular flow rate might not represent a large impact on the overall efficiency of the propulsor, as the losses due intake duct and nozzle will comprise separate and independent considerations, quite independent of the efficiency of the pump design.

@@

\subsection{Modern literature and results}
Abundant modern works, including those which involve experimental validation of theoretical results also clearly confirm that restrictions on the efficiency of waterjets are enduring, and give us a clear starting-point for consideration of their probable orders of magnitude, and likely transition points indicating at what speeds pumpjets might become advantageously more efficient.  It should be noted however, that relatively little literature focuses on the performance of propulsive systems far far outside of their intended design speeds, which tends to be far higher than the patrol and transit speeds of diesel electric submarines.  Fewer still tend to express their work in terms of ship-speed, instead adopting the convention of using dimensionless ratios so that work done on one system can be readily translated into different combinations loads, pressures, speeds and sizes.  However, from a few notable studies exist which provide important reference points. (Further interpretation of the what we can lean from the trends expressed in dimensionless coefficients will be discussed later.)

@@

Both of these previous studies confirm the general and widespread assumption that jets are known to be more efficient at high speeds, and propellers at lower speeds.  They also suggest that differences in efficiency could be very large, of a factor of two or more, including at speeds well over a submarine's transit speed.  However, care must be taken making precise quantitative comparisons, since such experiments involved jets which expelled into air (waterjets) as opposed to remaining under water (pumpjets) do have at least one significant difference in their makeup, though their internal pump mechanisms may often be similar. Specifically, pumpjets would experience drag on the exterior of the shroud/duct, which waterjets would not, however waterjets are likely to have longer intake ducts, which might involve larger duct losses than pumpjets.  Given that these factors might affect efficiency, and the overall design of the nozzle considerably, more direct comparison with fully submerged jets should be sought to make more direct comparisons.

At IMARESTS 10th International Naval Engineering Conference, a paper was presented by BMT Defence and Rolls Royce \parencite{giles2010}, outlining a proposal for 'fully submerged waterjets' to be used on future surface ships in order to achieve better acoustic performance, particularly for an anti-submarine warfare role. The jets comprise a mixed-flow pump coupled to an electric motor, and have been both extensively modeled and also tested in demonstrator vessels.  These may be a closer representation of the performance characteristics achievable by submerged pumpjets, since shape of the intake and shroud more closely resembles the shape of a pumpjet that might be found on a submarine.  Importantly, this work focuses specifically on the performance of the jets as a function of ship-speed, including identifying possible crossover points.

@@

The paper makes no effort to justify the selection of choice of jets on the efficiency of the propulsion system, instead focusing on improved acoustic performance, and potential machinery space-saving advantages.  The vessel under consideration would require 250 tonnes more fuel with jets find than with propellers, the jets would require more power to reach equal water-speed to a conventionally propelled craft right up to 30 kt.  In the case of this analysis, modifications to the hull-form in order to accommodate the jet (which are likely to be more substantial for a surface ship than a submarine) also have an impact on the resistance offered by the hull.  This also has an interaction with the different shape of the propulsors, which alters the required level of delivered (effective) power to achieve a given speed, in addition to the considerations of how efficiently the propulsor generates that effective power. The paper states: "This is an encouraging result for a first iteration as the AWJ-21™ arrangement has not been fully optimised and, hence, there is further prospect of reducing the shaft power for this form. If the power could be reduced by 5\% (e.g. by fur ther improving the waterjet/hull fairing), then the AWJ-21™ vessel would be better at all speeds above 25 knots."  This statement, however, is made in reference to the hull-form needed to accommodate the jet, rather than the jet propulsion efficiency itself \parencite{giles2010}.  As the propulsive efficiency curves given in the study show, the propulsors themselves are considerably further from reaching parity, still separated by about 10\% at 25kt.

@@

Moreover, recent literature seems to confirm that efficiency of pumpjets fall off precipitously towards zero the closer that the vessel speed comes to zero.  This strongly suggests the existence of some fundamental constraint in the physics of the operation of jets, which recent advances in technology seems not to have overcome.  To more firmly establish the extent to which adaptation or plausible improvements in technology could or could not change this trend, it is worthwhile considering the theoretical limits that are presented by the physics.  This should further limit the range of error that might be made in estimating realistic bounds of performance for a system which we do not have precise design specifications for.

@@

\begin{figure}[h]
\captionsetup{width=.8\textwidth}
\includegraphics[width=.8\textwidth, center]{JetLowFlow.png}
\caption{At low flow rates the water has insufficient velocity to overcome the adverse pressure gradients created by the duct shape, and flow separation begins to occur on the duct intake, and between the blades beginning near the roots (stall).  Flow through the jet would be unsteady, with excessive energy expended creating vortices and turbulence that results in little thrust.}
\label{fig:JetLowFlow.png}
\end{figure}
